Increased productivity using Atomify - a real-time LAMMPS visualizer

### 2018-06-13 IOPLaTeX.tex
Switched to IOP

\documentclass[aps,pre,twocolumn,letterpaper,floatfix,nofootinbib]{revtex4}
	% for \code color
\usepackage[dvipsnames]{xcolor}
\usepackage{graphicx}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{pdfpages}
\usepackage{afterpage}
\usepackage[hidelinks]{hyperref}
\usepackage{epstopdf}
\usepackage{todonotes}
\usepackage{menukeys}
\usepackage{siunitx}
\usepackage{listings}
% Pretty formating of c++ (see \cpp command)
\usepackage{xspace}
\newcommand*{\cpp}{C\ensuremath{++}\xspace}

@@

\lstset{
numbers=left,
numberstyle=\small,
numbersep=8pt,
frame = single,
language=LAMMPS,
framexleftmargin=15pt,
xleftmargin=0.65cm,
xrightmargin=0.1cm}

@@

\title{Increased productivity using Atomify - a real-time LAMMPS visualizer}
\author{Anders Hafreager$^1$}
\author{Svenn-Arne Dragly$^{1}$}
%\author{Anders Malthe-S\o renssen$^1$}
\affiliation{$^1$Department of Physics - University of Oslo\\Sem S{\ae}lands vei 24, NO-0316, Oslo, Norway }
\date{\today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
%
The typical workflow when running atomistic simulations includes working with several programs.
A text editor is needed to create and modify the scripts, the terminal to run the simulation, and programs like VMD or Ovito to visualize the system over time.
If physical quantities are computed, the data is often plotted using programming languages like MATLAB or Python,
where additional scripts must be used.
This is a tedious process, especially for teaching purposes and for people who are new in the field.
We here introduce Atomify; a high performance real-time visualizer for atomistic simulations that can simulate and render more than one million atoms with excellent frame rate on modern hardware.
Atomify supports OpenMP acceleration, GPU acceleration, real-time plotting of physical quantities, and an easy-to-use code editor in one single application.
It currently uses LAMMPS as physics engine, but it can be extended to support other codes like Gromacs, NAMD or OpenMM.
Atomify is open-source software (GPL) written in C++ using the Qt framework.
%
\end{abstract}

\maketitle

@@

Over the past decades, users of LAMMPS\citep{Plimpton1995Fast} and other molecular dynamics simulators such as GROMACS, NAMD and CHARMM\citep{berendsen1995gromacs, Phillips2005Scalable, brooks2009charmm} have
typically written scripts in a text editor and run the scripts by passing them to the simulator
on the command line.
The simulator in turn writes the results to disk and the user analyzes the results
by reading the files manually or passing them to other programs.
With the arrival of software like VMD,\citep{Humphrey1996Vmd}
it become easy to both visualize the trajectories and certain physical quantities.
VMD came with support both for reading the results from file or attaching to a
running process of supported simulators.
Later, modern visualization tools like Ovito\citep{Stukowski2009Visualization} has made both visualization and analysis more
accessible to new users through its powerful scriptable modifier pipeline.
In addition, the results can be loaded into common programming languages like MATLAB and Python
to produce publication-ready figures.

In combination, the text editors, the simulators, the visualization software, and the programming
languages like MATLAB and Python define the workflow of a typical researcher in computational
materials science.
While this is a powerful workflow in well-established simulations,
it is tedious for prototyping new ideas.
The workflow requires a fairly large amount of context switching,
moving from the text editor to the terminal to run the simulation,
waiting for the results, and then finally being able to study them using the analysis
and visualization tools.

@@

It is especially useful during development of new scripts, but can be used
to run full, publication-ready simulations with minimal simulation overhead.

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{gui.png}
	\caption{%
    Overview of the Graphical User Interface (GUI) in Atomify.
    The toolbar (1) is used to change between the different modes, such as
    editing, viewing and analyzing the running simulation,
    in addition to playback controls.
    The file browser (2) contain the list of open LAMMPS scripts or data files.
    The text editor (3) provides code highlighting.
    The viewport (4) shows the current simulation.
    The simulation properties (5) shows the simulations details, gives access to
    LAMMPS objects such as groups, regions and computes, and the rendering
    properties (6) lets you change the viewport settings, such as lighting and
    draw mode.
    TODO: Add numbers to figure.
    }
	\label{fig:gui}
\end{figure*}

@@

\section{\label{sec:features}Summary of features}
Atomify was designed to improve workflow with reduced number of
context switches and mouse clicks to obtain immediate results from a simulation.
Its main features can be summarized as script editing,
real-time visualization and plotting of physical quantities.
This turns out to improve the workflow substantially.
In addition, Atomify comes with multiple examples,
many of which have been made by members of the LAMMPS community.

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{figures/rightbar.pdf}
	\caption{
		The \textit{Simulation summary} is a powerful feature in Atomify where relevant
		properties from LAMMPS is presented real-time.
		In a), a list of all groups defined is shown with the current atom count.
		b) shows all defined regions. Each group and region can be hovered so that atoms
		are colored after what group / region they are in. If an atom is in multiple
		groups or regions, they will get the color of the latest one defined.
		The color boxes can be clicked which will either always highlight the atoms in a certain group or region,
		or if clicked again, hide all of them.
		In c) and d), a list of all variables and computes is shown. If a variable or compute
		produces a scalar, the current value is shown as can be seen with all the variables.
		By clicking the name, a plot will show up showing the time evolution of that quantity.
		If a per-atom variable or compute is hovered, all atoms will be colored according to their
		current scalar value and a histogram over these values is shown if clicked.
		The result of hovering atoms defined by the groups in our case study is shown in Fig. \ref{fig:cylinder_simulation}.
    }
	\label{fig:rightbar}
\end{figure}

It provides a lightweight script editor with syntax highlighting and line numbers.
Multiple files can be open at the same time and the state persist across working sessions.
The script can easily be restarted by clicking the button in the GUI or using
the default shortcut (\keys{Ctrl+R} in Windows and Linux or \keys{\cmd+R} in macOS).
This enables the user to get immediate visual feedback on changes in the script.

\section{\label{sec:casestudy}Case study: flow through a narrow nanotube}
The features of Atomify is best demonstrated through a case study where many of
them are useful to obtain a productive workflow during script development.
We have chosen to create a system with fluid flow inside a tight nanochannel.
This involves creating a somewhat complicated initial geometry with multiple regions
where we apply different rules such as a dissipative thermostat, constant force mimicking
applied pressure gradient and frozen atoms to avoid net momentum of the solid.

Fluid flow in porous media is often characterized with its permeability - the inverse flow resistance of a material.
The permeability can be measured using Darcy's law when applying a pressure gradient.
In Darcy's law, the permeability is assumed to only depend on the material and its geometry.
However, this turns out not being the case for dilute
gases where the measured permeability changes as a function of pressure\citep{klinkenberg1941permeability}.
This effect is called the Klinkenberg effect.

The discrepancy is explained by the fact that at low pressures, the fluid velocity
at the boundary is non-zero, a non-slip boundary condition.
A non-slip boundary condition is a good approximation for liquids in larger pores,
but if the channel size $R$ is of the same order,
or larger, as the mean free path $\lambda$ of the fluid, this is no longer true.
The ratio between the mean free path and the channel size is quantified through
the Knudsen number, $\text{Kn} = \lambda / R$, and can be used to apply the Klinkenberg correction\citep{klinkenberg1941permeability}
to correctly describe the \textit{effective permeability} as a function of pressure in the high Knudsen number limit.

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{lj_flow/configuration.png}
	\caption{
		Initial configuration in our case study simulation.
		A periodic box consisting of $(40\times25\times25)$ FCC unit cells with reduced density $\rho^* = 0.8442$ where we will have gas flow in a cylinder with radius 10 (reduced length).
		The different colors indicate the regions of which different integration rules apply.
		Blue atoms are fixed and do not move at all,
		whereas the yellow atoms are thermalized with a Langevin\citep{schneider1978molecular} dissipative thermostat to mimick the heat exchange with the frozen bulk atoms.
		The red and the green atoms are integrated normally, except the green gas atoms have an applied constant acceleration representing the pressure gradient.
		Setting up and verifying the system is simple using Atomify due to i.e. the real-time coloring of groups and regions.
    }
	\label{fig:cylinder_simulation}
\end{figure}

In this case study, we will use Atomify to measure the velocity profile of a
Lennard Jones gas inside a nanotube with varying density to observe
the slip velocity effect at low densities.

We create the nanotube inside a Lennard Jones solid of size  $(40\times25\times25)$ unit cells with reduced density $\rho^* = 0.8442$.
The fluid in the nanochannel, or cylinder, will flow due to an induced pressure gradient, see Fig. \ref{fig:cylinder_simulation}.
Here we apply a pressure gradient $\nabla P$ in the $x$-direction on the gas atoms using a constant acceleration $a$ calculated as
\begin{align}
	a = \frac{\nabla P}{\rho_m},
\end{align}
where $\rho_m$ is the mass density of the gas atoms.

Applying such an acceleration means adding energy to the system which in turn will increase the temperature and eventually melt the system.
To avoid this, we will couple the solid to an external heat bath so the added energy can be transferred out from the system through a dissipative thermostat.

Even though we only apply a pressure gradient on the gas atoms, the induced momentum will also be transferred to the solid atoms which eventually will start moving in the positive $x$-direction which is not desired.
We therefore freeze the outmost atoms (blue in Fig \ref{fig:cylinder_simulation}).
To obtain a system like this, we divide the atoms into multiple regions, which are integrated differently:

\begin{itemize}
	\item $r \leq 10$: applied pressure gradient
	\item $r \leq 17$: regular time integration
	\item $14 < r \leq 17$: Dissipative Langevin thermostat
	\item $17 < r$: frozen.
\end{itemize}
The different radii are specified in reduced Lennard Jones units.

\subsection{Creating the initial geometry}
We create a new script in Atomify using \keys{\cmd+N}, and save it (\keys{\cmd+S}) as \code{run.in}.
Start from the following input script
\lstinputlisting[language=LAMMPS]{simple.in}
we will modify it part by part.
Here, we have a Lennard Jones solid with reduced density
$\rho^* = 0.8442$, reduced mass $m^* = 1.0$ with $(40\times25\times25)$
FCC unit cells and a cutoff of $2.5\sigma$ where $\sigma$ is 1.0 in the reduced units.

To make sure LAMMPS executes our modifiction before visualization is started,
we need to make sure that additional commands happen \textit{before} the \code{run 1000} command.
If not, LAMMPS will run 1000 timesteps before getting to our modifications.

To define the inner cylinder, we first specify the radius and the center before using the region command
\begin{lstlisting}[basicstyle=\tiny, frame = none, numbers=none, framexleftmargin=0pt, xleftmargin=-0.75cm, xrightmargin=0.0cm]
	variable R equal $(ly*0.25)
	variable c equal $(ly*0.5)
	region cylinder1 cylinder x $c $c $R EDGE EDGE
\end{lstlisting}
The region command will create a cylinder $C_1$ in the $x$-direction, with $(y,z)$-coordinates at the system center \code{\$c} with radius \code{\$R} using the full length of the cylinder specified when using \code{EDGE}.

We now press \keys{\cmd+R} to run the simulation to see how it looks.
As shown in Fig. \ref{fig:rightbar}, we can hover the specified region to highlight the atoms in that region.

When we hover the region, we immediately see that the cylinder is in fact not placed at the center of the system, as shown in Fig. \ref{fig:initial_configuration}a.
The reason is that LAMMPS operate with two different units, either the length unit or the lattice length unit. This is explained in the documentation, but is a common mistake.
Discovering mistakes like this is very easy using Atomify.
The problem here is that LAMMPS expects the values to be in \textit{lattice} units, i.e. multiples of the lattice constant.
Since we used \code{\$(0.5*lx)} - which is in the actual lenght units - we need to specify \code{units box} in the region command.
We then rerun the simulation with \keys{\cmd+R} which gives the expected result in figure \ref{fig:initial_configuration}b.

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{figures/initial_configuration.pdf}
	\caption{
		Several snapshots of the process of creating the simulation in our case study.
		This shows how one typically generate a geometry step by step where the immediate
		feedback is very useful to quickly discover mistakes.
		The wrongly placed cylinder is shown in a) whereas it is placed correctly in b).
		c) shows how the system looks after deleting atoms to obtain $\rho = 0.01$. The next cylinder $C_2$ is shown in d),
		and all regions (placed in groups) are shown in e). Here we have uniquely defined the purple shell which should have a dissipative thermostat.
		In f), the list of all groups defined in this system is shown with colors matching those of e). All figures are rendered with Atomify.
    }
	\label{fig:initial_configuration}
\end{figure}
We now want to delete atoms so that we have a certain density. The volume of the inner cylinder is $\pi R^2 L$ which can be calculated in LAMMPS with \code{variable V equal PI*v\_R*v\_R*lx}.
We have chosen reduced mass units so the mass of our atoms is 1. The number density $\rho$ then equals the mass density $\rho_m$.
The command \code{delete\_atoms} can be used: \code{delete\_atoms porosity region-ID fraction seed}, where we need to specify the fraction of the existing atoms in that region we want to delete, and a random seed which can be any positive number.
Since we want to specify a certain density, we need to figure out how many atoms we should delete which depends on the current number of atoms inside the region.

To count the number of atoms inside a region, we first create a group containing all the atoms in it, and then create a variable that counts them.
Assuming density $\rho = 0.1$, we can calculate the expected number inside the cylinder as $N = V\rho$.
The atoms inside the region can now be deleted with the following commands
\begin{lstlisting}[basicstyle=\tiny, frame = none, numbers=none, framexleftmargin=0pt, xleftmargin=-0.75cm, xrightmargin=0.0cm]
	group cylinder1 region cylinder1
	variable N_cyl equal count(cylinder1)
	variable V equal PI*v_R*v_R*lx
	variable rho equal 0.1
	variable N_wanted equal $V*${rho}
	variable delete_fraction equal (${N_cyl}-${N_wanted})/${N_cyl}
	delete_atoms porosity cylinder1 ${delete_fraction} 1234
\end{lstlisting}
and the result is shown after rerunning again with \keys{\cmd+R}, see Fig. \ref{fig:initial_configuration}c.

This is a powerful workflow with Atomify where quick changes can be tested with immediate feedback.
Each time we add or change a command, we just rerun with \keys{\cmd+R} to see how it looks like.

The next step now is to create a larger cylinder $C_2$, the one containing the red atoms in Fig. \ref{fig:cylinder_simulation}, where regular time integration will take place.
This cylinder should have a slightly bigger radius and can be created with another region command:
\begin{lstlisting}[basicstyle=\tiny, frame = none, numbers=none, framexleftmargin=0pt, xleftmargin=-0.75cm, xrightmargin=0.0cm]
	region cylinder2 cylinder x $c $c $(v_R+4) EDGE EDGE
\end{lstlisting}
Again press \keys{\cmd+R} and see the expected result in Fig. \ref{fig:initial_configuration}d.

Now we have the inner cylinder with the gas atoms defined as a region and the next cylinder shell (including the gas atoms) defined as another region.
We now want to create another shell, the one with the dissipative thermostat.
All the atoms with the dissipative thermostat, the inner shell and the gas will be integrated with the \code{fix nve} integrator.
We therefore create a larger cylinder $C_3$ with $r=17$ which will be the one defining the group for \code{fix nve}.
With all three regions defined, we can also create groups which can be used in fixes
\begin{lstlisting}[basicstyle=\tiny, frame = none, numbers=none, framexleftmargin=0pt, xleftmargin=-0.75cm, xrightmargin=0.0cm]
	region cylinder3 cylinder x $c $c $(v_R+7) EDGE EDGE
	group cylinder1 region cylinder1
	group cylinder2 region cylinder2
	group cylinder3 region cylinder3
	group nve region cylinder3
\end{lstlisting}
where we also have created the group \code{nve} for clarity.

The dissipative thermostat will be \code{fix langevin}\citep{schneider1978molecular}, but it should only be applied on the outer shell defined by atoms being in $C_3$, but not in $C_2$.
Groups define sets on which we can apply regular set operations like union, intersection and subtraction.
We can obtain the outer shell with the command

\begin{lstlisting}[basicstyle=\tiny, frame = none, numbers=none, framexleftmargin=0pt, xleftmargin=-0.75cm, xrightmargin=0.0cm]
	group langevin subtract cylinder3 cylinder2
	group frozen subtract all cylinder3
	group gas dynamic all region cylinder1 every 1
\end{lstlisting}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%novalidate

\documentclass[12pt,a4paper,final]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
\usepackage{iopams}
\usepackage{graphicx}
\usepackage[breaklinks=true,colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}

% Custom for this paper
\bibliographystyle{iopart-num}
%\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{mathtools}
% for \code color
\usepackage[dvipsnames]{xcolor}
\usepackage{todonotes}
\usepackage{menukeys}
\usepackage{siunitx}
\usepackage{listings}
% Pretty formating of c++ (see \cpp command)
\usepackage{xspace}
\newcommand*{\cpp}{C\ensuremath{++}\xspace}

@@

\lstset{
basicstyle=\small,
frame = single,
language=LAMMPS,
framexleftmargin=15pt,
xleftmargin=0.65cm,
xrightmargin=0.1cm}

@@

\title{Increased productivity using Atomify - a real-time LAMMPS visualizer}

\author[cor1]{Anders Hafreager$^{1}$}
\address{$^1$Department of Physics - University of Oslo\\Sem S{\ae}lands vei 24, NO-0316, Oslo, Norway}
\eads{\mailto{anderhaf@fys.uio.no}, \mailto{andershaf@gmail.com}}

\author{Svenn-Arne Dragly$^{1}$}
\ead{s.a.dragly@fys.uio.no}

\begin{abstract}
The typical workflow when running atomistic simulations includes working with several programs.
A text editor is needed to create and modify the scripts, the terminal to run the simulation, and programs like VMD or Ovito to visualize the system over time.
If physical quantities are computed, the data is often plotted using programming languages like MATLAB or Python,
where additional scripts must be used.
This is a tedious process, especially for teaching purposes and for people who are new in the field.
We here introduce Atomify; a high performance real-time visualizer for atomistic simulations that can simulate and render more than one million atoms with excellent frame rate on modern hardware.
Atomify supports OpenMP acceleration, GPU acceleration, real-time plotting of physical quantities, and an easy-to-use code editor in one single application.
It currently uses LAMMPS as physics engine, but it can be extended to support other codes like Gromacs, NAMD or OpenMM.
Atomify is open-source software (GPL) written in C++ using the Qt framework.
\end{abstract}

%Uncomment for PACS numbers title message
\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB? 
\vspace{2pc}
\noindent{\it Keywords}: Article preparation, IOP journals
% Uncomment for Submitted to journal title message
\submitto{\JPA}
% Comment out if separate title page not required
%\maketitle

@@

Over the past decades, users of LAMMPS\cite{Plimpton1995Fast} and other molecular dynamics simulators such as GROMACS, NAMD and CHARMM\cite{berendsen1995gromacs, Phillips2005Scalable, brooks2009charmm} have
typically written scripts in a text editor and run the scripts by passing them to the simulator
on the command line.
The simulator in turn writes the results to disk and the user analyzes the results
by reading the files manually or passing them to other programs.
With the arrival of software like VMD,\cite{Humphrey1996Vmd}
it became easy to both visualize the trajectories and certain 
physical quantities such as the radial distribution function.
VMD came with support both for reading the results from file or attaching to a
running process of supported simulators.
\todo[inline]{Why is Ovito more modern than VMD?}
Later, modern visualization tools like Ovito\cite{Stukowski2009Visualization} has made both visualization and analysis more
accessible to new users through its powerful scriptable modifier pipeline.
In addition, the results can be loaded into common programming languages like MATLAB and Python
to produce publication-ready figures.

In combination, the text editors, the simulators, the visualization software, and the programming
languages like MATLAB and Python define the workflow of a typical researcher in computational
materials science.
While this is a powerful workflow in well-established simulations,
it is tedious for prototyping new ideas.
This workflow requires a fairly large amount of context switching,
moving from the text editor to the terminal to run the simulation,
waiting for the results, and then finally being able to study them using the analysis
and visualization tools.

@@

It is especially useful during development of new scripts, but can be used
to run full, publication-ready simulations with minimal simulation overhead.
The current implementation only supports LAMMPS, but is written in a general
way so integrating other codes is a simple task.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gui.png}
	\caption{%
    Overview of the Graphical User Interface (GUI) in Atomify.
    The toolbar (1) is used to change between the different modes, such as
    editing, viewing and analyzing the running simulation,
    in addition to playback controls.
    The file browser (2) contain the list of open LAMMPS scripts or data files.
    The text editor (3) provides code highlighting.
    The viewport (4) shows the current simulation.
    The simulation properties (5) shows the simulations details, gives access to
    LAMMPS objects such as groups, regions and computes, and the rendering
    properties (6) lets you change the viewport settings, such as lighting and
    draw mode.
    TODO: Add numbers to figure.
    }
	\label{fig:gui}
\end{figure}

@@

\begin{figure}[htp!]
	\centering
	\includegraphics[width=0.5\textwidth]{figures/rightbar.pdf}
	\caption{
		The \textit{Simulation summary} is a powerful feature in Atomify where relevant
		properties from LAMMPS is presented real-time.
		In a), a list of all groups defined is shown with the current atom count.
		b) shows all defined regions. Each group and region can be hovered so that atoms
		are colored after what group / region they are in. If an atom is in multiple
		groups or regions, they will get the color of the latest one defined.
		The color boxes can be clicked which will either always highlight the atoms in a certain group or region,
		or if clicked again, hide all of them.
		In c) and d), a list of all variables and computes is shown. If a variable or compute
		produces a scalar, the current value is shown as can be seen with all the variables.
		By clicking the name, a plot will show up showing the time evolution of that quantity.
		If a per-atom variable or compute is hovered, all atoms will be colored according to their
		current scalar value and a histogram over these values is shown if clicked.
		The result of hovering atoms defined by the groups in our case study is shown in figure \ref{fig:cylinder_simulation}.
    }
	\label{fig:rightbar}
\end{figure}

\section{\label{sec:features}Summary of features}
Atomify was designed to improve the workflow by reducing the number of
context switches and mouse clicks to obtain immediate results from a simulation.
It has a script editor, visualization window and a summary panel so all relevant parts
of a simulation can be on the screen at the same time.

Any changes to the script like defining regions or groups,
testing temperatures, size of simulation box or inserting atoms can
quickly be verified since visualization is real-time.
Rerunning the current script can be done by pressing \keys{Ctrl+R} (\keys{cmd+R} on macOS).

How the pressure, radial distribution function or some other
complicated variable evolves over time is easily understood
since these can be plotted real-time with a simple click.
Bonds between atoms in a molecule is automatically extracted from
the internals of LAMMPS, or can be defined as a distance threshold
between atom types.

The Atomify GUI is shown in figure \ref{fig:gui} where \todo{short summary of GUI}.

This turns out to improve the workflow substantially.
In addition, Atomify comes with multiple examples,
many of which have been made by members of the LAMMPS community.

\section{\label{sec:casestudy}Case study: flow through a narrow nanotube}
The features of Atomify is best demonstrated through a case study where many of
them are useful to obtain a productive workflow during script development.
We have chosen to create a system with fluid flow inside a tight nanochannel.
This involves creating a somewhat complicated initial geometry with multiple regions
where we apply different rules such as a dissipative thermostat, constant force mimicking
applied pressure gradient, and frozen atoms to avoid net momentum of the solid.
We will create two different atom types so we can have a gas phase and a solid phase at the same temperature.

Fluid flow in porous media is often characterized with its permeability - the inverse flow resistance of a material.
The permeability can be measured using Darcy's law when applying a pressure gradient $\nabla P$.
In Darcy's law, the permeability is assumed to only depend on the material and its geometry.
However, this turns out not being the case for dilute
gases where the measured permeability changes as a function of pressure\cite{klinkenberg1941permeability}.
This is called the Klinkenberg effect.

The discrepancy is explained by the fact that at low pressures, the fluid velocity
at the boundary is non-zero, i.e. we have a non-slip boundary condition.
A non-slip boundary condition is a good approximation for liquids in larger pores,
but if the channel size $R$ is of the same order,
or larger, as the mean free path $\lambda$ of the fluid, it is no longer true.

The ratio between the mean free path and the channel size is quantified through
the Knudsen number, $\text{Kn} = \lambda / R$, and can be used to apply the Klinkenberg correction\cite{klinkenberg1941permeability}
to correctly describe the \textit{effective permeability} as a function of pressure in the high Knudsen number limit.

\begin{figure}
	\centering
	\includegraphics[width=0.75\textwidth]{lj_flow/configuration.png}
	\caption{
		Initial configuration in our case study simulation.
		A periodic box consisting of $(40\times25\times25)$ FCC unit cells with reduced density $\rho^* = 0.8442$ where we will have gas flow in a cylinder with radius 10 (reduced length).
		The different colors indicate the regions of which different integration rules apply.
		Cyan atoms are fixed and do not move at all,
		whereas the purple atoms are thermalized with a Langevin\cite{schneider1978molecular} dissipative thermostat to mimick the heat exchange with the frozen bulk atoms.
		The orange and the yellow atoms are integrated normally, but the yellow gas atoms also have an applied constant acceleration representing the pressure gradient.
		Setting up and verifying the system is simple using Atomify due to i.e. the real-time coloring of groups and regions.
    }
	\label{fig:cylinder_simulation}
\end{figure}

In this case study, we will use Atomify to measure the velocity profile of a
gas inside a nanotube with varying density to observe
the slip velocity effect at low densities.
We will use the Lennard Jones potential with two different atom types.

We create the nanotube inside a solid with  $(60\times30\times30)$ FCC unit cells and reduced density $\rho^* = 0.975$.
This density was chosen so that the two atom types with the Lennard Jones coefficients specified below will coexist in different phases.
The fluid in the nanotube, or cylinder, will flow due to an induced pressure gradient, see figure \ref{fig:cylinder_simulation}.
Here we apply a pressure gradient $\nabla P$ in the $x$-direction on the gas atoms using a constant acceleration $a$ calculated as
\[
    a = \frac{\nabla P}{\rho_m},
\]
where $\rho_m$ is the mass density of the gas atoms and $\nabla P$ is the desired pressure gradient.
This represents a pressure difference $\Delta P$ over a small volume element $\Delta x$ on a cross-sectional area $\Delta y\Delta z$.

Applying an acceleration means adding energy to the system which in turn will increase the temperature and eventually melt the system.
To avoid this, we couple the solid to an external heat bath so the added energy can be transferred out from the system through a dissipative thermostat.

Even though we only apply a pressure gradient on the gas atoms, the induced momentum will be transferred
to the solid atoms which eventually will start moving in the positive $x$-direction, which is not desired.
The solid should be at rest since it is only the relative velocity between the gas and the solid that defines
the flow. We therefore freeze the outmost atoms (colored cyan in Fig \ref{fig:cylinder_simulation}).
To obtain this system, we divide the atoms into multiple cylindrical regions, which are integrated differently:

\begin{itemize}
	\item $r \leq 12$: applied pressure gradient
	\item $r \leq 23$: regular time integration
	\item $18 < r \leq 23$: dissipative Langevin thermostat
	\item $23 < r$: frozen.
\end{itemize}
The different radii are specified in reduced Lennard Jones units.

\subsection{Creating the initial geometry}
We create a new script in Atomify using \keys{\cmd+N}, and save it (\keys{\cmd+S}) as \code{run.in}.
Copy the following input script

\todo[inline]{Should we remove line numbers?}
\lstinputlisting[language=LAMMPS]{simple.in}
we will modify it part by part.
Here, we have a Lennard Jones solid with reduced density
$\rho^* = 0.975$, reduced masses $m_1^* = 4.0$, $m_2^* = 0.5$ with $(60\times30\times30)$
FCC unit cells and a cutoff of $2.5$. 
The force field parameters are chosen so that atoms of type 1
are in the solid phase with type 2 atoms being gas at reduced temperature $T^*=0.5$.
The two atom types are weakly coupled (small $\epsilon_{ij}$)so the gas atoms don't
stick too much to the surface which would prevent flow.

To make sure LAMMPS executes our modifiction before visualization is started,
we need to make sure that following additional commands happen \textit{before} the \code{run 1000} command.
If not, LAMMPS will run 1000 timesteps before getting to our changes.

To define the inner cylinder, we first specify the radius and the center before using the region command
\begin{lstlisting}
variable R equal 12
variable c equal $(ly*0.5)
region cylinder1 cylinder x $c $c $R EDGE EDGE
set region cylinder1 type 2
\end{lstlisting}
The region command will create a cylinder $C_1$ in the $x$-direction, with $(y,z)$-coordinates at the system center \code{\$c} with radius \code{\$R} using the full length of the system in the $x$-direction specified when using \code{EDGE}.
Atoms inside the region will also be changed to type 2 since these will be the gas atoms.

We now press \keys{\cmd+R} to run the simulation to see how it looks.
As shown in figure \ref{fig:rightbar}, we can hover the specified region to highlight the atoms in it.

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{figures/initial_configuration.pdf}
	\caption{
		Several snapshots of the process of creating the simulation in our case study.
		This shows how one typically generate a geometry step by step where the immediate
		feedback is very useful to quickly discover mistakes.
		The wrongly placed cylinder is shown in a) whereas it is placed correctly in b).
		c) shows how the system looks after deleting atoms to obtain $\rho = 0.01$. The next cylinder $C_2$ is shown in d),
		and all regions (placed in groups) are shown in e). Here we have uniquely defined the purple shell which should have a dissipative thermostat.
		In f), the list of all groups defined in this system is shown with colors matching those of e). All figures are rendered with Atomify.
    }
	\label{fig:initial_configuration}
\end{figure}

When we hover the region, we immediately see that the cylinder is in fact not placed at the center of the system, as shown in figure \ref{fig:initial_configuration}a.
The reason is that LAMMPS operate with two different units, either the length unit or the lattice length unit. This is explained in the documentation, but is a common mistake.
Discovering mistakes like this is very easy using Atomify.

The problem here is that LAMMPS expects the values to be in \textit{lattice} units, i.e. multiples of the lattice constant.
Since we used \code{\$(0.5*lx)}, which gives the system length in actual lenght units, we need to specify \code{units box} in the region command.
We then rerun the simulation with \keys{\cmd+R} which gives the expected result in figure \ref{fig:initial_configuration}b.

Now that the region looks like we expect, we want to delete atoms to control the density.
The volume of the inner cylinder is $\pi R^2 L$ which can be calculated in LAMMPS with \code{variable V equal PI*v\_R*v\_R*lx}.
The command \code{delete\_atoms} can be used: \code{delete\_atoms porosity region-ID fraction seed},
where we need to specify the fraction of the existing atoms in that region we want to delete, and a random seed which can be any positive number.
Since we want to specify a certain density, we need to figure out how many atoms we
should delete which depends on the current number of atoms inside the region.

To count the number of atoms inside a region, we first create a group containing all the atoms in it, and then create a variable that counts them.
Assuming density $\rho_m = 0.1$, we can calculate the expected number inside the cylinder as $N = V\rho_m/m$,
where $V$ is the volume of the cylinder and $m$ is the mass of type 2 atoms.
The atoms inside the region can now be deleted with the following commands
\begin{lstlisting}
group cylinder1 region cylinder1
variable N_cyl equal count(cylinder1)
variable V equal PI*v_R*v_R*lx
variable rho_m equal 0.1
variable N_wanted equal $V*${rho_m}/0.5
variable delete_fraction equal (${N_cyl}-${N_wanted})/${N_cyl}
delete_atoms porosity cylinder1 ${delete_fraction} 1234
\end{lstlisting}
where we see the result by rerunning with \keys{\cmd+R} in figure \ref{fig:initial_configuration}c.

This is a powerful workflow with Atomify where changes quickly can be tested with immediate feedback.
Each time we add or change a command, we just rerun with \keys{\cmd+R} to see how it looks like.

The next step now is to create a larger cylinder $C_2$, the one containing the orange atoms in
figure \ref{fig:cylinder_simulation}, where regular time integration will take place.
This cylinder should have a slightly bigger radius and can be created with another region command:
\begin{lstlisting}
region cylinder2 cylinder x $c $c $(v_R+6) EDGE EDGE units box
\end{lstlisting}
Again press \keys{\cmd+R} to see the expected result in figure \ref{fig:initial_configuration}d.

Now we have the inner cylinder with the gas atoms defined as a region and the next cylinder shell (including the gas atoms) defined as another region.
We now want to create another shell, the one with the dissipative thermostat.
All the atoms with the dissipative thermostat, the inner shell and the gas will be integrated with the \code{fix nve} integrator.
We therefore create a larger cylinder $C_3$ with $r=23$ which will be the one defining the group for \code{fix nve}.
With all three regions defined, we also create groups which will be used in fixes
\begin{lstlisting}
region cylinder3 cylinder x $c $c $(v_R+11) EDGE EDGE units box
group cylinder1 region cylinder1
group cylinder2 region cylinder2
group cylinder3 region cylinder3
group nve region cylinder3
\end{lstlisting}
where we also have created the group \code{nve} for clarity.

We will use \code{fix langevin}\cite{schneider1978molecular} as the dissipative thermostat,
but it should only be applied on the outer shell defined by atoms being in $C_3$, but not in $C_2$.
Groups define sets on which we can apply regular set operations like union, intersection and subtraction.
We can obtain the outer shell with the command

\begin{lstlisting}
group langevin subtract cylinder3 cylinder2
group frozen subtract all cylinder3
group gas dynamic all region cylinder1 every 1
\end{lstlisting}
